### 2016-03-21 doc/publications/plos16/plos16.tex
plos paper comments

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Article Title - Submission to PLOS Journals}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Name1 Surname\textsuperscript{1,\Yinyang},
Name2 Surname\textsuperscript{2,\Yinyang},
Name3 Surname\textsuperscript{2,\textcurrency a},
Name4 Surname\textsuperscript{2,\ddag},
Name5 Surname\textsuperscript{2,\ddag},
Name6 Surname\textsuperscript{2},
Name7 Surname\textsuperscript{3,*},
with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{2} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{3} Affiliation Dept/Program/Center, Institution Name, City, State, Country
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
\ddag These authors also contributed equally to this work.

% Current address notes
\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
\dag Deceased

% Group/Consortium Author Note
\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* CorrespondingAuthor@institute.edu

@@

In the past few years, whole genome sequence (WGS) data has rapidly been becoming an essential part in the clinical space, where traditional methods are failing to produce results.

WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],

One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
types of data

Developed by Apache, Hadoop is one platform for dealing with large data sets is 
Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data. Spark, based on Scala, allows for distributed applications that can scale to hundreds or thousands of nodes.
A Spark application runs on a `driver' node, and this node divides and distributes work out to the many `worker' nodes.
Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries. Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as
unsupervised approaches such as k-means clustering.
We previously demonstrated the versatility and scalability of Spark by developing VariantSpark, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework. Using VariantSpark,
we successfully built a k-means model on the 2500 individuals from the 1000 Genomes Project, based on their genomic data. We also investigated classifying the samples using Random Forests. We chose random forests due to its
ability to capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise overfitting and variance.

However, we observe that Spark ML's random forests implementation isn't able to handle the 1000 Genomes data, and cannot build a model on the entire dataset. Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2),
the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model. This means that the work doesn't scale as while the driver is processing data, the worker nodes are lying unused.

The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
for nodes on the computer cluster to handle.

Therefore, we develop our own 'wide' version of these ML algorithms. Unlike the standard Apache implementation, these are able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.

%sit~\cite{bib1} Nullam Eq.~(\ref{eq:schemeP}) ~\cite{bib2}  Table~\ref{table1} CO\textsubscript{2}  see \nameref{S1_Text}.
%\begin{equation}\label{eq:schemeP} 
%D_{coll} = \frac{D_f+\frac{[S]^2}{K_D S_T} D_S} {1+\frac{[S]^2}{K_D S_T}}, 
%D_{sm} = \frac{D_f+ \frac{[S]}{K_D} D_S}{1+\frac{[S]}{K_D}},
%\end{equation}

@@

\subsection*{Machine learning with Spark ML}
In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar.

There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.

One benefit is that hyperparameter optimisation is built in. Although possible in

\subsection*{CursedForest copes with the curse of dimensionality}
With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
These

\subsection*{Third section}

>>>>

% Title must be 250 characters or less.
% Please capitalize all terms in the title except conjunctions, prepositions, and articles.
\begin{flushleft}
{\Large
\textbf\newline{Cursed Forest - A random forest implementation for 'big' and 'wide' data}
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1},
Piotr Szul\textsuperscript{2},
Stephanie Li\textsuperscript{3},
James Doecke\textsuperscript{3},
Nick Ellis\textsuperscript{4},
Robert Dunne\textsuperscript{5}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\bf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\bf{2} Data61, CSIRO, Brisbane, QLD, Australia
\\
\bf{3} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\bf{4} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\\
\bf{5} Data61, CSIRO, Sydney, NSW, Australia
\\
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* Denis.Bauer@CSIRO.au

@@

The digital revolution or 'datafication' sees a dramatic increase in data collected about every aspect of life~\cite{Loebbecke2015}. 
These dataset not only grow vertically by capturing more and more events but also horizontally by capturing more information about these events.
The challenge of big and 'wide' data is especially pronounced in the health space where whole genome sequence (WGS) technology enabled researchers to interrogate all 3 billion basepairs of the human genome. 
While a large fraction of this data is identical between individuals, identifying the relevant and disease specific differences is the focus of complex trait analysis [citation].

For example, the 1000 Genome data consists of approximately 2500 samples with up to 80 million variants~\cite{1KG2012}, denoting the differences between individuals. 
This dataset consumes close to a terabyte of disk space.
The richness of this data can be utilised in other datasets as well by filling in or "imputing" the information at genomic positions previously unobserved due to the less detailed resolution of older array technology [citation about imputation]. 
This means that the large number of samples for which genome-wide association (GWA) data is available~\cite{Welter2013} can be extended horizontally by imputing the previously unobserved genotypes. 
This provides the potential to generate datasets of hundreds of thousands of individuals with millions of variants, highlighting the need for incorporating modern compute paradigms to deal with these challenges.

%WGS has proven itself in discovering rare variants [http://www.sciencedirect.com/science/article/pii/S2352396414000498] and diagnosing unknown genetic disorders [MATT MIGHT],
%One challenge with analysing WGS data over, for example, traditional genome-wide association (GWA) data, is developing software and algorithms to deal with the huge amounts of data.
%Although the sample size of WGS data-sets may be comparable to GWA, the dimensionality of the data may be many times larger. For example, the 1000 Genome data consists of
%approximately 2500 individuals with up to 80 million variants. The uncompressed file sizes are close to a terabyte. Of course, it's not just genomic data that is rapidly growing. Many other
%types of data

%Developed by Apache, Hadoop is one platform for dealing with large data sets is 
%Platforms Fortunately computational tools such as Apache Spark are available as a framework for iteratively dealing with big data. 
Distributing the compute tasks is one strategy to overcome this challenge and Apache Spark, based on Scala, allows for distributed applications to scale to hundreds or thousands of nodes.
A Spark application runs on a `driver' node, and this node divides and distributes work out to the many `worker' nodes.
Spark also includes the machine learning (ML) library, Spark ML, providing popular machine learning libraries. 
Spark ML includes supervised approaches such as decision trees and the ensemble method, random forests, as well as unsupervised approaches such as k-means clustering.

We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a framework allowing users to easily analyse Variant Call Format (VCF) files using ML algorithms on the Spark framework. 
Using VariantSpark, we successfully built a k-means model on the full 2500x80Million matrix to cluster individuals by their ethnicity achieving an Adjusted Rand Index ARI=0.84, with 1 perfect and -1 random clustering.

While unsupervised clustering is useful for selected clinical applications~\cite{Li2015}, a more common analysis task is supervised learning, i.e. given a phenotype of clinical label determine the most associated features. 
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary by fitting a linear combination of the feature variables.
Processing the 'wide' genomic data poses difficulties for machine learning applications as having more features than samples causes the model to not generalise well and overfit the data. 
This is known as the 'curse of dimensionality'~\cite{Bauer2014}. 
Random Forests (RF) as an ensemble of weak learners overcomes this issue and is hence ideally suited to this application. 
Furthermore, RF are able to capture interactions between features, which is of importance for modelling complex polygenic diseases.

%We also investigated classifying the samples using Random Forests. 
%We chose random forests due to its
%ability to capture complex interactions between features. Also, due to it being an ensemble method where trees can be formed independently, which, is ideal for parallelisation. Furthermore, the ensemble of weak learners can help to minimise overfitting and variance.

However, we observe that Spark's standard RF implementation is not able to handle the extremely 'wide' genomic data as it was developed for large number of samples with only modest dimensionality. 
Although Spark ML can build a RF model on a subset of the data (chromosome 1~and~2), the time taken is excessive due to the huge amount of data being aggregated and processed by the single driver node during intermediate stages of building the model (see Result~\ref{comp}). 
This unbalanced work load where the driver node becomes the bottleneck and worker nodes being idle prevents a seamless scaling to larger datasets.

%The reason for the failure can be attributed to the "curse of dimensionality". Where Spark ML is designed to process a huge number of samples with a modest dimensionality, the 1000 Genomes data
%instead contains a huge number of dimensions (features). As a result, the default random forests implementation is unable to partition the data into small-enough tasks
%for nodes on the computer cluster to handle.

We therefore we developed CursedForest a Spark-based random forest algorithm able to process big and 'wide' data. Unlike the standard Apache implementation, CursedForest is able to easily process data with millions of dimensions without excessive resource requirements on the driver or worker nodes.
%Also, due to more efficient handling of high-dimensional data, equivalent jobs are likely to finish faster on wideML than Spark ML.
In the fist section we

@@

%OPTIONS
% Spark ML vs MLlib
% random forest, kmeans, logistic regression 
% synthetic data demonstrating random forest can cope with n<<p
% James dataset
% GWAS dataset

\subsection*{Synthetic data demonstrating that RF can deal with extreme dimensionality}
Detailing random association between features and labels
and whatever rob did

\subsection*{CursedForest copes with the curse of dimensionality}
\label{comp}
Spark ML vs CursedForest vs Kmeans vs LR on small and full version (where applicable) of 1000G (performance and time/mem)

%In this section we compare Apache's higher level machine learning library, Spark ML, with MLLib, the standard low-level library. From the results, we see that times are similar. 
%There are many practical benefits for using Spark ML with VariantSpark, with a selection of these benefits derived from the `pipeline' system Spark ML uses.
%One benefit is that hyperparameter optimisation is built in. Although possible in

%\subsection*{CursedForest copes with the curse of dimensionality}
%With Spark ML unable to cope with high-dimensional data, we develop a `wide' implementations of these ML algorithms. **technical stuff by Piotr here**
%These

\subsection*{James' dataset}
Comparing CursedForest with feature selection + linear model (performance and time/mem)
